---
output: 
  stevetemplates::article:
    fig_caption: true
  
bibliography: master.bib
biblio-style: apsr
title: "Constructing a Structural Equation Model for Predicting Biological Age: A Technical Report under $AMAˆ{-1}$ perspective"
thanks: "Replication files are available on the author's Github account (https://github.com/AlvaroNovillo). **Current version**: `r format(Sys.time(), '%B %d, %Y')`; **Corresponding author**: alvanovi@ucm.es."
author:
- name: Álvaro Novillo Correas
  affiliation: 100483441@alumnos.uc3m.es
abstract: "This article discusses the implementation of $AMA^{-1}$, a methodology that emphasizes the idea that statistics is used to solve real-world problems. The acronym $AMA^{-1}$, represents the three fundamental steps in this methodology: understanding the real-world problem and related data (A), translating the real-world problem into a statistical problem and solving it (M), and translating the statistical results back into terms of the original problem ($A^{-1}$). The article presents basic aspects of each step along with an example of its implementation"
keywords: "$AMA^{-1}$, SEM"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
# spacing: double
endnote: no

---

```{r setup, include=FALSE}
library(sjPlot)
library(ggplot2)  
library(kableExtra)
library(parameters)
knitr::opts_chunk$set(cache=TRUE,
                      message=FALSE, warning=FALSE,
                      fig.path='figs/',
                      cache.path = '_cache/',
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })
```


# Introduction (A)
In this article, we will implement the AMA-1 methodology in R, dealing with the challenge of estimating biological age using observed health indicators. The dataset^[The dataset can be found at: https://aulaglobal.uc3m.es/mod/resource/view.php?id=4711061] comprises information on weight, sickness status, and chronological age. To do so, we will follow the three fundamental steps of the methodology: understanding the real-world problem and related data (A), translating the real-world problem into a statistical problem and solving it (M), and translating the statistical results back into terms of the original problem (A-1).

```{r summary, echo=FALSE, fig.cap= "Histogram of the variables contained in the dataset"}
# Step A: Understanding the real-world problem and related data
# Read the dataset
data <- read.csv("bioage.csv", sep=";")
colnames(data)[2] <- "sick"



# Get summary statistics for each variable
#summary(data) %>%
#  kbl(booktabs = T, caption = "Summary of the dataset") %>%
#  kable_styling(position = "center")

# Create histograms of the variables
#par(mfrow = c(1, 3))  # Set up a 1x3 layout for the histograms

#hist(data$weight, main = "Weight Histogram", xlab = "Weight")
#hist(data$sick, main = "sick Histogram", xlab = "sick")
#hist(data$age, main = "Age Histogram", xlab = "Age")

```

Making a preliminary analysis of the dataset, we can see that over half of the animals in our sample are affected by illness, with an illness status average of 0.6. Additionally, the age variable is left-skewed, suggesting a higher frequency of middle-aged animals. 


# Method (M)
Employing Structural Equation Modeling (SEM)^[The lavaan package in R is employed for model estimation using the Maximum Likelihood (ML) method. The algorithm iteratively refines parameter estimates to maximize the likelihood of observing the given data under the specified model. See @lavaan for further insight.], we construct a model to predict biological age as a latent variable. The relationship between the latent variable and each observed variable is quantified by the factor loadings. Standardization is applied to the observed variables to facilitate meaningful interpretation of coefficients and improve numerical stability in the estimation process.

Next, to make a comparison between biological age and measured chronological age, a regression analysis will be conducted using obtained biological age and chronological age values.  If the regression coefficient is found to be significant and approaches one, it would demonstrate the correlation between the latent and observed variables, indicating that the animals are undergoing regular and healthy aging processes.

```{r}
model <- '
  # Measurement model
    bio_age =~ weight + sick

  # Regression model
    age ~ bio_age
'

```

```{r, echo = FALSE}
# Step M: Translating the real-world problem into a statistical problem and solving it
# Fit a linear regression model to the data
#model <- lm(age ~ ., data=data)
# Load the required package
# install.packages("lavaan")
library(lavaan)

# Standardize the variables
df_standardized <- data.frame(
  weight = scale(data$weight),
  sick = scale(data$sick),
  age = scale(data$age)
)

# Create a model with standardized variables
model <- '
  # Measurement model
    bio_age =~ weight + sick

  # Regression model
    age ~ bio_age
'

# Fit the model
fit <- sem(model, data = df_standardized)


# Get the latent variable values (biological age)
latent_vars <- as.data.frame(fitted(fit, "lv"))




#summary(fit,standardized=TRUE, fit.measures=TRUE)
mp <- model_parameters(fit)
print_md(mp)
```

Based on these results, given the factor loading (Table. 1) for weight and sick we can conclude that they are a significant predictor of biological age in these animals. Furthermore, the regression coefficient (Table. 2) of 0.934 indicates a positive correlation between biological age and chronological age, confirming a direct relationship between the latent and observed variables. Given the $95\%$ confidence interval calculated for this coefficient, it is not possible to make any claims regarding whether the animals' chronological age corresponds to their biological age. To do so, it would be necessary to narrow the confidence interval, which could be achieved by increasing the sample size of the population.

# Results ($A^{-1}$)

The constructed Structural Equation Model (SEM) serves as a valuable framework for unraveling the intricate connections between observed health indicators and the latent construct of biological age. The model, supported by strong factor loadings and a significant regression coefficient, underscores the pivotal role of sickness status and weight in capturing the essence of biological aging.

While the model exhibits commendable fit indices, acknowledging its limitations is imperative for a comprehensive interpretation. The current sample size, while sufficient for initial exploration, beckons for expansion to enhance robustness and generalizability. A larger and more diverse dataset would fortify the model's capacity to discern nuanced patterns within the aging process.

Furthermore, recognizing the influence of contextual factors is crucial. Incorporating a richer contextual understanding of the database could empower us to address the intricacies of biological age estimation with a more sophisticated statistical approach.


# References
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\vspace*{-0.2in}
\noindent







